---
title: "ae_table"
format: html
editor: visual
---

# AE summary {#sec-aesummary}


Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we summarize number of participants that were included in each safety analysis in Section 12.2, Adverse Events (AEs).

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(tidyr) # Manipulate data
library(gt)
# library(r2rtf) # Reporting in RTF format
```


The data used to summarize AE information is in `adsl` and `adae` datasets.

```{r}
adsl <- read_sas("https://github.com/elong0527/r4csr/raw/main/data-adam/adsl.sas7bdat")
adae <- read_sas("https://github.com/elong0527/r4csr/raw/main/data-adam/adae.sas7bdat")
```

We first summarize participants in population by treatment arm.

```{r}
pop <- adsl %>%
  filter(SAFFL == "Y") %>%
  rename(TRTAN = TRT01AN) %>%
  count(TRTAN, name = "tot")

pop
```

We transform the data to simplify the analysis of each required AE criteria of interest.

- With one or more adverse events
- With drug-related adverse events
- With serious adverse events
- With serious drug-related adverse events
- Who died

```{r}
tidy_ae <- adae %>%
  mutate(
    all = SAFFL == "Y",
    drug = AEREL %in% c("POSSIBLE", "PROBABLE"),
    ser = AESER == "Y",
    drug_ser = drug & ser,
    die = AEOUT == "FATAL"
  ) %>%
  select(USUBJID, TRTAN, all, drug, ser, drug_ser, die) %>%
  pivot_longer(cols = c(all, drug, ser, drug_ser, die))

tidy_ae %>% head(4)
```

We summarize the number and percentage of participants who meet each AE criteria.

```{r}
fmt_num <- function(x, digits, width = digits + 4) {
  formatC(
    x,
    digits = digits,
    format = "f",
    width = width
  )
}
```

```{r}
ana <- tidy_ae %>%
  filter(value == TRUE) %>%
  group_by(TRTAN, name) %>%
  summarise(n = n_distinct(USUBJID)) %>%
  left_join(pop, by = "TRTAN") %>%
  mutate(
    pct = fmt_num(n / tot * 100, digits = 1),
    n = fmt_num(n, digits = 0),
    pct = paste0("(", pct, ")")
  )

ana %>% head(4)
```

We prepare reporting-ready dataset for each AE group.

```{r}
t_ae <- ana %>%
  pivot_wider(
    id_cols = "name",
    names_from = TRTAN,
    values_from = c(n, pct),
    values_fill = list(
      n = "   0",
      pct = "(  0.0)"
    )
  )

t_ae <- t_ae %>%
  mutate(name = factor(
    name,
    c("all", "drug", "ser", "drug_ser", "die"),
    c(
      "With one or more adverse events",
      "With drug-related adverse events",
      "With serious adverse events",
      "With serious drug-related adverse events",
      "Who died"
    )
  )) %>%
  arrange(name)
```

We prepare reporting-ready dataset for the analysis population.

```{r}
t_pop <- pop %>%
  mutate(
    name = "Participants in population",
    tot = fmt_num(tot, digits = 0)
  ) %>%
  pivot_wider(
    id_cols = name,
    names_from = TRTAN,
    names_prefix = "n_",
    values_from = tot
  )

t_pop
```

The final report data is saved in `tbl_ae_summary`.

```{r}
tbl_ae_summary <- bind_rows(t_pop, t_ae) %>%
  select(name, ends_with("_0"), ends_with("_54"), ends_with("_81"))

tbl_ae_summary
```



## Analysis of Adverse Event Summary


```{r}
gt_tbl <-
  tbl_ae_summary %>%
  gt(rowname_col = "name") %>%
  tab_header(
    title = "Analysis of Adverse Event Summary",
    subtitle = "(Safety Analysis Population)"
  ) %>%
  cols_label(
    starts_with("n") ~ "n",
    starts_with("pct") ~ "(%)"
  ) %>%
  tab_spanner(
    columns = ends_with("0"),
    label = "Placebo"
  ) %>%
  tab_spanner(
    columns = ends_with("54"),
    label = "Xanomeline Low Dose"
  ) %>%
  tab_spanner(
    columns = ends_with("81"),
    label = "Xanomeline High Dose"
  ) %>%
  sub_missing(missing_text = "") %>%
  cols_width(
    stub() ~ px(300),
    everything() ~ px(90)
  ) %>%
  tab_footnote(footnote = "Every subject is counted a single time for each applicable row and column.") %>%
  tab_options(
    page.orientation = "landscape",
    page.numbering = TRUE,
    page.header.use_tbl_headings = TRUE,
    page.footer.use_tbl_notes = TRUE
  )
  
gt_tbl
```
Writing the table to HTML can be done with `gtsave()`.

```{r}
gt_tbl %>% gtsave("Analysis of Adverse Event Summary.html")
```

Write the **gt** table to an RTF document.

```{r}
gt_tbl %>% gtsave("Analysis of Adverse Event Summary.rtf")
```






